Modelling the travel time of transit vehicles in \rt through a GTFS-based road network using GPS vehicle locations

### 2020-03-31 anzjs2018/anzjs.tex
Add line numbers
Editor changes

\def\volumeyear{2018}
% \linenumbers

@@

\begin{abstract}
Predicting the arrival time of a transit vehicle involves not only
knowledge of its current position and schedule adherence,
but also traffic conditions along the remainder of the route.
Road networks are dynamic 
and can quickly change from free-flowing to highly congested,
which impacts the arrival time of transit vehicles,
particularly buses which often share the road with other vehicles,
so reliable predictions need to account for \rt and future traffic conditions.
The first step in this process is to construct a framework with which road state
(traffic conditions) can be estimated using \rt transit vehicle position data.
Our proposed framework implements a vehicle model using a particle filter
to estimate road travel times,
which are used in a second model to estimate \rt traffic conditions.
Although development and testing took place in Auckland, New Zealand,
we generalised each component to make the framework compatible 
with other public transport systems around the world.
We demonstrate the \rt feasibility and performance of our approach in real-time,
where a combination of R and C++ was used to obtain the necessary performance results.
Future work will use these estimated traffic conditions 
in combination with historical data to obtain reliable
arrival time predictions of transit vehicles.
\end{abstract}

>>>>

\def\volumeyear{2018}
\linenumbers

@@

\begin{abstract}
Predicting the arrival time of a transit vehicle involves not only
knowledge of its current position and schedule adherence,
but also traffic conditions along the remainder of the route.
Road networks are dynamic
and can quickly change from free-flowing to highly congested,
which impacts the arrival time of transit vehicles,
particularly buses which often share the road with other vehicles,
so reliable predictions need to account for \rt and future traffic conditions.
The first step in this process is to construct a framework with which road state
(traffic conditions) can be estimated using \rt transit vehicle position data.
Our proposed framework implements a vehicle model using a particle filter
to estimate road travel times,
which are used in a second model to estimate \rt traffic conditions.
Although development and testing took place in Auckland, New Zealand,
we generalised each component to make the framework compatible
with other public transport systems around the world.
We demonstrate the \rt feasibility and performance of our approach in real-time,
where a combination of \textsf{R} and \textsf{C++} was used to obtain the necessary performance results.
Future work will use these estimated traffic conditions
in combination with historical data to obtain reliable
arrival time predictions of transit vehicles.
\end{abstract}

### 2020-03-31 anzjs2018/sections/01_introduction.tex
Editor changes

In public transport, \rt information (RTI)---%
most notably estimated times of arrival (ETAs)---%
keeps transport users informed of changes to their commute,
allowing them to plan their journey accordingly.
Previous research has shown that perceived waiting time is less
when arrival time information is available \citep{TCRP_2003b};
however, in Auckland, RTI is highly unreliable,
leading to frustration and ultimately deterring public transport users.
Inaccurate ETAs are the primary source of this frustration,
as they fluctuate---sometimes dramatically---over time.
While unpredictable changes to traffic conditions may cause this,
it can equally be due to poor schedule calibration.
Furthermore, buses are shown as \emph{on-time} 
if they are not present in the \rt system and 
the service provider has not manually cancelled them,
in which case ETAs are based solely on scheduled arrival times.
Once the scheduled arrival time has passed,
the service is removed from the \rt board,
leaving passengers unsure as to when---if at all---their bus will show up,
a phenomenon referred to as ``ghost buses'' by transport bloggers.

>>>>

In public transport, \rt information (RTI)---%
most notably estimated times of arrival (ETAs)---%
keeps transport users informed of changes to their commute,
allowing them to plan their journey accordingly.
Previous research has shown that perceived waiting time is less
when arrival time information is available \citep{TCRP_2003b};
however, in Auckland, RTI is highly unreliable,
leading to frustration and ultimately deterring public transport users.
Inaccurate ETAs are the primary source of this frustration,
as they fluctuate---sometimes dramatically---over time.
While unpredictable changes to traffic conditions may cause this,
it can equally be due to poor schedule calibration.
Furthermore, buses are shown as \emph{on-time}
if they are not present in the \rt system and
the service provider has not manually cancelled them,
in which case ETAs are based solely on scheduled arrival times.
Once the scheduled arrival time has passed,
the service is removed from the \rt board,
leaving passengers unsure as to when---if at all---their bus will show up,
a phenomenon referred to as `ghost buses' by transport bloggers.

### 2020-03-31 anzjs2018/sections/02_gtfs.tex
Editor changes

GTFS (general transit feed specification)
is an API (application programming interface) specification for transit data
detailing how it should be organised,
making access easier for application developers.
GTFS, developed and maintained by Google \citep{GoogleDevelopers_2006},
% who use it in Google Maps Transit Directions,
is used by over 1100~transit providers around the world\footnote{%
source: \url{http://transitfeeds.com} as of August 2019},
including here in Auckland, New Zealand.
An advantage of this standardised format is that,
provided an application depends solely on GTFS data,
after developing it locally in Auckland it will be deployable to any other GTFS-based
public transport system with minimal modification.

There are two components to GTFS.
The first, \emph{GTFS static}, includes information about
\begin{itemize}
\item \emph{stops}, a physical location where passengers can embark and disembark the vehicle;
\item \emph{routes}, a sequence of two or more stops displayed as a single service;
\item \emph{trips}, an instance of a route occurring at a specific time of day;
\item \emph{schedules}, specifying the arrival (and departure) times for each bus at each of its stops; 
\item and \emph{shapes}, the sequence of points defining a vehicle's path along a route.
\end{itemize}
The second component, \emph{GTFS-realtime},
is only available in a subset of providers due to the requirement of 
on-board GPS tracking devices and a central server.
It provides a standardised format for sharing vehicle positions and trip delays,
which developers access via an API for use in \rt applications.

>>>>

GTFS (general transit feed specification)
is an API (application programming interface) specification for transit data
detailing how it should be organised,
making access easier for application developers.
GTFS, developed and maintained by Google \citep{GoogleDevelopers_2006},
% who use it in Google Maps Transit Directions,
is used by over 1100~transit providers around the world
(sourced from \url{http://transitfeeds.com} on August 2019),
including here in Auckland, New Zealand.
An advantage of this standardised format is that,
provided an application depends solely on GTFS data,
after developing it locally in Auckland it will be deployable to any other GTFS-based
public transport system with minimal modification.

There are two components to GTFS.
The first, \emph{GTFS static}, includes information about:
\begin{itemize}
\item \emph{stops}, a physical location where passengers can embark and disembark the vehicle;
\item \emph{routes}, a sequence of two or more stops displayed as a single service;
\item \emph{trips}, an instance of a route occurring at a specific time of day;
\item \emph{schedules}, specifying the arrival (and departure) times for each bus at each of its stops;
\item and \emph{shapes}, the sequence of points defining a vehicle's path along a route.
\end{itemize}
The second component, \emph{GTFS-realtime},
is only available in a subset of providers due to the requirement of
on-board GPS tracking devices and a central server.
It provides a standardised format for sharing vehicle positions and trip delays,
which developers access via an API for use in \rt applications.

### 2020-03-31 anzjs2018/sections/04_implementation.tex
Add line numbers
minor edits
Editor changes

The application consists of two components:
the first handles importing static GTFS data and constructing the network,
while the second implements the \rt models and prediction.
We used Rcpp \citep{Rcpp}
to develop the program,
which provides access to R \citep{rcore},
which contains many useful tools for data manipulation and package development,
as well as the speed and memory management capabilities of C++.
The program is implemented in the R package
\verb+transitr+, available on Github (\url{https://github.com/tmelliott/transitr}).
In this section, we discuss the features of the \rt component
and assess its performance 
with respect to timing and travel time estimation.

Below is the general structure of the \rt component, 
where the bold steps are those discussed in this paper.
\begin{enumerate}
\item Load GTFS data from database
\item Each time new data are received \ldots
\begin{enumerate}
    \item Update or create new vehicle objects from the new data
    \item \textbf{Run particle filter on each vehicle to update or initialise its state}
    \item \textbf{Update state of any roads for which vehicles 
        have completed travel}
    \item Generate ETAs for vehicles using combination of vehicle and network state
    \item Write ETAs to a file for distribution
\end{enumerate}
\end{enumerate}

During the development of the application,
our primary concern was to ensure each component of the program
was as efficient as possible,
so ETA generation and distribution is fast enough to be feasible in real-time,
with a target of 30~seconds or faster at peak time.
Using C++ provided the memory management control necessary to make this possible.
Since we modelled each vehicle independently, 
it was straightforward to spread the vehicle updates 
over multiple cores using OpenMP without thread-safety concerns.

The number of particles needed per vehicle 
depends on many factors,
so it was necessary to explore the performance of the application
with a varying number of particles.
We assessed the application's performance
using a range of values for
system noise and GPS error.
To enable comparisons, we implemented a simulated \rt environment
in which the same subset of real vehicle data from 8~October 2018
could be analysed using a range of settings.
These simulations were carried out on a virtual machine 
with 8~Intel Xeon 3.00GHz CPU cores and 32~GB of memory, 
running Ubuntu~16.04 and using R~3.4.1.

@@

The most time-consuming component was ETA writing,
which involved summarising the individual ETAs estimated for each particle.
While not discussed here this involved computing quantiles which uses a sorting algorithm
with complexity $O(N \log N)$,
which explains the non-linear relationship with $N$.

@@

The last measure of performance is the relative variance of segment travel times.
We used relative variance as each segment has a very different travel time distribution,
so to compare the variability between simulations we needed to 
compute the average variance of estimated travel times over all simulations,
taking the average (over road segments) of the ratio between variance for
each simulation and variance over all simulations.
If $\tilde \bz_\ell^{\omega,\sigma}$ is a vector of all travel times
along segment $\ell$ during the simulation with GPS error $\omega$ and system noise $\sigma$,
then the relative variance $\bar v^{\omega,\sigma}$ is calculated by
\begin{equation*}
\begin{split}
v_\ell^{\omega,\sigma} &=
\frac{\mathrm{Var}(\tilde \bz_\ell^{\omega,\sigma})}{\mathrm{Var}(\tilde \bz_\ell)} \\
\bar v^{\omega,\sigma} &=
    \frac{1}{L}\sum_{\ell=1}^L v_\ell^{\omega,\sigma}
\end{split}.
\end{equation*}
Figure~\ref{fig:travel_times} shows that simulations with larger GPS error
resulted in a higher relative variance,
while simulations with more particles had a smaller relative variance.
There was no clear pattern to the effect of system noise.

>>>>

The application consists of two components:
the first handles importing static GTFS data and constructing the network,
while the second implements the \rt models and prediction.
We used \textsf{Rcpp} \citep{Rcpp}
to develop the program,
which provides access to \textsf{R} \citep{rcore},
which contains many useful tools for data manipulation and package development,
as well as the speed and memory management capabilities of \textsf{C++}.
The program is implemented in the \textsf{R} package
\verb+transitr+, available on Github (\url{https://github.com/tmelliott/transitr}).
In this section, we discuss the features of the \rt component
and assess its performance
with respect to timing and travel time estimation.

Below is the general structure of the \rt component,
where the bold steps are those discussed in this paper:
\begin{enumerate}
\item Load GTFS data from database
\item Each time new data are received, do:
\begin{enumerate}
    \item Update or create new vehicle objects from the new data
    \item \textbf{Run particle filter on each vehicle to update or initialise its state}
    \item \textbf{Update state of any roads for which vehicles
        have completed travel}
    \item Generate ETAs for vehicles using combination of vehicle and network state
    \item Write ETAs to a file for distribution
\end{enumerate}
\end{enumerate}

During the development of the application,
our primary concern was to ensure each component of the program
was as efficient as possible,
so ETA generation and distribution is fast enough to be feasible in real-time,
with a target of 30~seconds or faster at peak time.
Using \textsf{C++} provided the memory management control necessary to make this possible.
Since we modelled each vehicle independently,
it was straightforward to spread the vehicle updates
over multiple cores using \textsf{OpenMP} without thread-safety concerns.

The number of particles needed per vehicle
depends on many factors,
so it was necessary to explore the performance of the application
with a varying number of particles.
We assessed the application's performance
using a range of values for
system noise and GPS error.
To enable comparisons, we implemented a simulated \rt environment
in which the same subset of real vehicle data from 8~October 2018
could be analysed using a range of settings.
These simulations were carried out on a virtual machine
with 8~Intel Xeon 3.00GHz CPU cores and 32~GB of memory,
running \textsf{Ubuntu}~16.04 and using \textsf{R}~3.4.1.
The simulations were processed locally using \textsf{R}~3.6.0,
and made use of the \textsf{R} packages \verb+dplyr+ \citep{dplyr}
for manipulating and summarising the results,
and \verb+ggplot2+ \citep{ggplot2} to graph them.

@@

The most time-consuming component was ETA writing,
which involved summarising the individual ETAs estimated for each particle.
While not discussed here this involved computing quantiles which uses a sorting algorithm
with complexity $\mathcal{O}(N \log N)$,
which explains the non-linear relationship with $N$.

@@

The last measure of performance is the relative variance of segment travel times.
We used relative variance as each segment has a very different travel time distribution,
so to compare the variability between simulations we needed to
compute the average variance of estimated travel times over all simulations,
taking the average (over road segments) of the ratio between variance for
each simulation and variance over all simulations.
If $\tilde \bz_\ell^{\omega,\sigma}$ is a vector of all travel times
along segment $\ell$ during the simulation with GPS error $\omega$ and system noise $\sigma$,
then the relative variance $\bar v^{\omega,\sigma}$ is calculated by
\begin{equation*}
\begin{split}
v_\ell^{\omega,\sigma} &=
\frac{\mathrm{Var}(\tilde \bz_\ell^{\omega,\sigma})}{\mathrm{Var}(\tilde \bz_\ell)}, \\
\bar v^{\omega,\sigma} &=
    \frac{1}{L}\sum_{\ell=1}^L v_\ell^{\omega,\sigma}.
\end{split}
\end{equation*}
Figure~\ref{fig:travel_times} shows that simulations with larger GPS error
resulted in a higher relative variance,
while simulations with more particles had a smaller relative variance.
There was no clear pattern to the effect of system noise.

### 2020-03-31 anzjs2018/sections/05_discuss.tex
Editor changes

\section{Discussion and Future Work}
\label{sec:discussion}

@@

The results of our simulations show that the framework is a 
feasible \rt alternative.
While the timings presented were for off-peak times,
further investigation into the choice of $N$ and $N_\text{eff}$,
as well as work removing redundancy from the transition algorithm,
should allow us to model all vehicles during peak periods---%
when there are 2-3~times as many vehicles---%
within our 30-second time frame target.

>>>>

\section{Discussion and future work}
\label{sec:discussion}

@@

The results of our simulations show that the framework is a
feasible \rt alternative.
While the timings presented were for off-peak times,
further investigation into the choice of $N$ and $N_\text{eff}$,
as well as work removing redundancy from the transition algorithm,
should allow us to model all vehicles during peak periods---%
when there are 2--3~times as many vehicles---%
within our 30-second time frame target.
